\section{Introduction}

Software often needs to respond to external events and data flows.
Consider for example software in interactive applications, for desktops,
web and mobile phones, in graphics and in processing sensor data from
phones or IoT-devices.  Traditionally, handling these asynchronous
events was done using the \emph{Observer design pattern}~\cite{johnson1995design}
or \emph{callback functions}~\cite{gallaba2015don}.  Using these
patterns, the system consuming the data does not have to block waiting
for methods to return, but instead receives a notification event when
data is available.  While these patterns decouple the consumer from the
producer of the data, they typically lead to dynamic registration, side
effects in the consumers, and inversion of control~\cite{salvaneschi2014empirical,edwards2009coherent}.

Reactive Programming (RP) is an alternative to these patterns for event
driven computation.  RP defines event streams as lazy collections and
provides operators that allow developers to deal with the complications
of asynchronous event handling.  It offers declarative and concise
syntax for composing streams of data, to express the complex reactive
behavior of these applications.  RP started in academia in the form of
Functional Reactive Programming (FRP)~\cite{elliott1997functional,elliott2009push,czaplicki2013asynchronous,maier2010deprecating,meyerovich2009flapjax},
but in recent years the use of RP has exploded.  Languages such as Elm~\cite
{czaplicki2012elm} and libraries such as Reactor~\cite{Gutierrez2017},
Akka~\cite{klangakka} and Rx~\cite{meijer2010subject} are being used by
companies like Netflix, Microsoft and Google, to build highly responsive
and scalable systems.  Front-end libraries like Angular\footnote{\url{https://angular.io/}},
that use RP in their foundations, are used by many large sites (9.1\% of Quantcast
Top 10k websites%
%URL to long for hbox; https://trends.builtwith.com/javascript/Angular-JS.
\footnote{\url{{https://trends.builtwith.com/}}, accessed 2017-06-20}).
Developers and companies alike attempted to standardize ``Reactive
Programming'' in the form of the Reactive Manifesto~\cite{boner2014reactive}.

While reactive programs might be more declarative and concise, RP does
not work well with traditional interactive debuggers, shipped with most
IDE\'s~\cite{salvaneschi2016debugging}.  RP borrows from Functional
Programming (FP) for its abstractions, its laziness and advocating the
use of ``pure'' lambda functions.  Those features contribute to a
control flow that is hidden inside the RP implementation library and
lead to non-linear execution of user code.  This results in not useful
stack traces, while breakpoints do not help either, since relevant variables
are frequently not in scope.  Furthermore, using a low level debugger makes
it harder to interact with the high level abstractions that RP provides.
Compared to imperative programming, there is limited scientific
knowledge regarding how developers debug reactive programs.  Traditional
imperative program debugging practices~\cite{beller2017behavior} do not
apply to RP~\cite{salvaneschi2016debugging}.

In this work we address the issue of RP debugging by designing and
implementing a high level debugger called RxFiddle for a popular version
of RP, namely Reactive Extensions (Rx).  RxFiddle (1) provides an
overview of the dependencies in the data flow, (2) enables a detailed
insight in the data flow and the timing of individual events, and (3)
allows developers to trace values back through the data flow.  To guide
our design we conducted interviews among professional developers.  After
building RxFiddle, we validated it with a user experiment.  The results
show that RxFiddle can help developers debug RP data flows faster.

To steer the research, we formulate the following research questions:

\begin{description}
    \item[RQ1]
        How do developers debug RP?  \\
        Before we design tools it is important to understand the
        problems arising in the the current state~\cite{singer2010examination}.
        Anecdotal evidence by a number of resources%
        \footnote{%
        \label{foot:contribdays}\url{http://contributordays.com/contributor-days/rxjs}}%
        \footnote{\url{https://staltz.com/how-to-debug-rxjs-code.html}}
        suggests that debugging RP is difficult.

    \item[RQ2]
        How can we design a tool that helps developers debug RP?  \\
        By examining the results of RQ1, the limitations of traditional
        debuggers and the opportunities that RP programs offer in terms
        of structure and explicit dependencies between data flows, we
        design a novel RP debugger.

    \item[RQ3]
        Can specialized RP debuggers speed up comprehension \&
        debugging?  \\
        To validate our design and examine whether specialized tooling
        can improve the experience we measure the speed and correctness
        of comprehension in an experiment.

\end{description}
