\section{Threats to validity}
\paragraph{External validity} For the interviews we selected 5
professional developers that were both available and worked on projects
involving RxJS.  The online experiment was open to anyone who wanted to
participate, and shared publicly.  These recruitment channels pose a
threat to generalizability:  different practices might exist in
different companies, different developer communities and for different
RP implementations \& languages.  Future work is needed on validating
the debugger in these different contexts.

Our code samples for the tasks are based on documentation samples and
common use cases for Rx; RxFiddle might perform different on actual
samples from practice, especially when the developer is familiar with
the project or domain.  The experiment consists of 2 small and 2 medium
tasks; for larger tasks the effect of using the debugger could be bigger
and therefore be better measurable.  Still, we chose for these smaller
tasks:  in the limited time of the subjects they could answer only so
many questions.  With the limited amount of time available, we still
show that a significant speed-up can be achieved in some cases.  We
leave it for future work to extend the experiment to include user code
\& larger systems.

\paragraph{Construct validity} We measure the time between the moment a
question is displayed and the moment its correct answer is submitted.
Even though our questions and code samples are short and were designed
to be read quickly, still some variation is introduced by different
reading speeds of subjects.  A setup where the question and code can be
read before the time is started can remedy this threat; but introduces
the problem of \emph{planning} when given unlimited time~\cite{ko2015practical}:
subjects can start planning their solution before the time starts.
Furthermore, subjects might have different strategies to validate their
(potentially correct) assumptions before submitting, ranging from going
over the answer once more, to immediately testing the answer by
submitting it.  However, explicitly stating that invalid answers do not
lead to penalty might introduce more guessing behavior.  Future studies
could use longer tasks, with preparation time to read the sample
software at hand, with a wizard-like experiment interface presenting one
short question at a time.

\paragraph{Internal validity} As a result of the recruitment method of
the experiment, a mixed group of developers took part, attracting even
those without Rx experience.  To reduce the variation in experience that
this introduces, we separately examined the results of more experienced
developers.

At the time of the experiment RxFiddle was already available online for
use, and furthermore some of the experiment subjects had already used
RxFiddle during piloting.  We mitigate this issue partially by providing
a instruction video at the start of the experiment, however subjects
with extensive experience with RxFiddle might bias the results.

The \emph{subject-expectancy effect}~\cite{ko2015practical} poses a
validity concern, since subjects who expect a certain outcome, may
behave in a way that ensures it.  Our subjects had the opportunity to
learn the context of the experiment and thus could be more motivated to
use RxFiddle than using the traditional debugger.  Our online experiment
captures motivation to some extend as drop-out (defined as quiting,
before having started all tasks) happens; the approximately equal
drop-out in both groups (RxFiddle 58.1\%, Console 58.5\%, Fisher's exact
test p-value $ 1 $), suggests no motivational differences.  Future
studies could offer subjects external motivation (e.g. by ranking
contenders and gamification~\cite{dicheva2015gamification} the
experiment, or organizing a raffle among top contenders), to limit the
threats introduced by motivation.

% \todo{look at \url{http://ieeexplore.ieee.org/document/7476654/\#full-text-section}}\\
% \todo{look at Contributors paper of Gousios}\\

% \textbf{Tasks.}

% However, we want RP debugging to be accessible for novice developers too; as the visualization techniques can be embedded in documentation.
